Translation of B Implementations to the LLVM: Control and Modularity

### 2013-09-09 translation.tex
Fixed issue #1. Tested code generation with multiple imports of the same machine.

Figure~\ref{tab:node-attr} summarizes the structure
of the abstract syntax that is relevant in the definition of the proposed
translation and introduces identifiers used in the definition of the
translation. For instance, and for the purpose of the translation, the relevant
components of a B implementation are: \B{id}, the name identifying the
implementation within the project; \B{see}, a set of references to other modules
containing definitions used in the implementation; \B{const}, a set of constant
names, their type and their values; \B{impo}, a sequence of optionally
prefixed B machines corresponding to external modules instantiated in the
implementation; \B{var}, a set of variable names, their type and possibly
additional functional restrictions forming the rest of the state; \B{init}, a
sequence of instructions executed upon initialization of the implementation;
\B{op}, a set of operations, each being an algorithmic description of the
different functionalities provided by the implementation.

@@

\paragraph{Call up} instructions apply B operations with some values and obtain
results. The LLVM code generator produces instructions to evaluate the operation
arguments, the address of the variables receiving the results as well as an
instance of the LLVM \llvm{call} instruction. In addition, the proposed
translation includes as argument the address of a structure storing the state of
the component corresponding to this operation. Note that this component may be
either the implementation being translated (in the case of a so-called local
operation), or an instance of an imported machine.
\begin{align*}
  \trad{c}{\B{Call}} \isdef 
  & \LET p_i \sep il = \trad{c.inp}{\ListOf{Inp}} \AND p_o \sep ol = 
  \trad{c.out}{\ListOf{Out}} \AND \\
  & \quad id = \Global(\B{c}) \AND ty = \llvm{"\PH{\state(\B{c.root})} *} \AND \\
  & \quad val \sep p = \IF \B{c.inst} = \text{None} \\
  & \quad \quad \THEN \llvm{"\%self\$"} \sep \llvm{""} \\
  & \quad \quad \ELSE \LET l = \lbl \IN \\
  & \quad \quad \quad l \sep \\
  & \quad \quad \quad \llvm{"\PH{l} = getelementptr \PH{ \state({\B{c.inst.root}})} \%self\$, i32 0, i32 \PH{\idx(\B{c.inst})} \nl"} \\
  & \IN \\
  & \quad \llvm{"\PH{p_i}} \\
  & \quad \llvm{\PH{p_o}} \\
  & \quad \llvm{\PH{p}} \\
  & \quad \llvm{call void \PH{id}(\PH{ty}* \PH{val}, \PH{il}, \PH{ol}) \nl"}
\end{align*}
The translation of input and output arguments produces two code snippets. The
first corresponds to the LLVM instruction sequence $p$ responsible for
evaluating the argument, and the second to the type $t$ and name $v$ of the
variable storing the resulting value:
\begin{align*}
  \trad{i}{Inp} \isdef 
  & \LET p \sep v \sep t = \trad{i}{\B{Expr}} \IN p \sep \llvm{"\PH{t} \PH{v}"}\\
  \trad{o}{Out} \isdef 
  & \LET p \sep v \sep t = \trad{o}{lv} \IN p \sep \llvm{"\PH{t} \PH{v}"}
\end{align*}

@@

Initialization is done is two steps: first all imported modules are initialized,
second the initialization clause is executed.
\begin{align*}
  \trad{i}{Init} \isdef
  & \llvm{"define void @\PH{\B{i.id}}\$init\$() \{ \nl"} \\
  & \quad\llvm{\PH{\trad{i.impl}{\ListOf{InitImp}}}} \\
  & \quad\llvm{\PH{\trad{i.init, "exit"}{\ListOf{\B{Inst}}}_L}} \\
  & \quad\llvm{exit: ret void \nl} \\
  & \llvm{\}"}
\end{align*}
The initialization of an imported module is performed by a call to the
corresponding initialization function:
\begin{align*}
  \lefteqn{\trad{m}{InitImp} \isdef } \\
  & \LET \ell = \lbl \IN \\
  & \quad \llvm{"\PH{\ell} = getelementptr \PH{\state(\B{m.root})}* \%self\$, i32 0, i32 \PH{\idx(\B{m})}\nl}\\
  & \quad \llvm{call void @\PH{\B{m.id}}\$init\$(\PH{\state(\B{m})}*
    \PH{\ell})"}
\end{align*}

@@

For each test file, indicates the B0 constructs found in the implementation.
\begin{center}
\newcommand{\no}[0]{--}
\newcommand{\yes}[0]{$\circ$}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Example nº & 01 & 02 & 03 & 04 & 05 & 06 & 07 \\
\hline
\hline
\multicolumn{8}{|c|}{clauses} \\
\hline
import w.o prefix & \no & \no & \no & \no & \no & \no & \yes \\
import w prefix & \no & \no & \no & \no & \no & \no & \no \\
variables & \yes & \yes & \yes & \yes & \yes & \yes & \no \\
constants & \no & \yes & \no & \no & \yes & \no & \yes \\
op. input & \yes & \no & \yes & \no & \no & \no & \no \\
op. output & \yes & \no & \yes & \no & \no & \no & \yes \\
\hline
\multicolumn{8}{|c|}{instructions} \\
\hline
block & \yes & \no & \yes & \yes & \no & \yes & \no \\
local var & \no & \no & \yes & \yes & \no & \yes & \yes \\
identity & \no & \no & \yes & \no & \no & \no & \no \\
becomes equal & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
operation call & \no & \no & \no & \no & \no & \no & \yes \\
condition & \yes & \no & \no & \no & \yes & \no & \yes \\
case & \no & \no & \no & \no & \no & \yes & \no \\
assertion & \no & \no & \no & \no & \no & \no & \no \\
sequence & \yes & \yes & \yes & \yes & \no & \yes & \yes \\
while & \no & \no & \no & \yes & \no & \no & \yes \\
\hline
\multicolumn{8}{|c|}{types} \\
\hline
INT & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
BOOL & \yes & \yes & \no & \no & \yes & \no & \yes \\
\hline
\multicolumn{8}{|c|}{expressions} \\
\hline
int lit & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
bool lit & \yes & \yes & \no & \yes & \yes & \no & \yes \\
vars & \yes & \yes & \yes & \yes & \no & \yes & \no \\
loc. vars & \no & \no & \yes & \yes & \no & \yes & \yes \\
const & \no & \yes & \no & \no & \yes & \no & \yes \\ 
succ & \no & \yes & \no & \yes & \yes & \no & \no \\
pred & \no & \yes & \no & \yes & \yes & \no & \no \\
+ & \yes & \no & \no & \yes & \no & \yes & \yes \\
- & \yes & \no & \no & \no & \no & \no & \yes \\
* & \no & \no & \no & \no & \no & \yes & \no \\
\hline
\multicolumn{8}{|c|}{formulas} \\
\hline
and & \no & \no & \no & \no & \yes & \no & \no \\
or & \no & \no & \no & \no & \yes & \no & \no \\
not & \no & \no & \no & \no & \yes & \no & \no \\
$<$ & \yes & \yes & \no & \no & \no & \no & \yes \\
$>$ & \no & \no & \no & \no & \yes & \no & \yes \\
\hline
\end{tabular}
\end{center}

>>>>

Figure~\ref{tab:node-attr} summarizes the structure
of the abstract syntax that is relevant in the definition of the proposed
translation and introduces identifiers used in the definition of the
translation. For instance, and for the purpose of the translation, the relevant
components of a B implementation are: \B{id}, the name identifying the
implementation within the project; \B{see}, a set of references to other modules
containing definitions used in the implementation; \B{const}, a set of constant
names, their type and their values; \B{impo}, a sequence of optionally
prefixed B machines corresponding to external modules instantiated in the
implementation; \B{var}, a set of variable names, their type and possibly
additional functional restrictions forming the rest of the state; \B{init}, a
sequence of instructions executed upon initialization of the implementation;
\B{op}, a set of operations, each being an algorithmic description of the
different functionalities provided by the implementation.\DD{Include the
possibility that the import is a machine without implementation. Instead it
has a list of typed, implementable, variables.}

@@

\paragraph{Call up} instructions apply B operations with some values and obtain
results. The LLVM code generator produces instructions to evaluate the operation
arguments, the address of the variables receiving the results as well as an
instance of the LLVM \llvm{call} instruction. In addition, the proposed
translation includes as argument the address of a structure storing the state of
the component corresponding to this operation. Note that this component may be
either the implementation being translated (in the case of a so-called local
operation), or an instance of an imported machine.
\begin{align*}
  \trad{c}{\B{Call}} \isdef 
  & \LET p_i \sep il = \trad{c.inp}{\ListOf{Inp}} \AND p_o \sep ol = 
  \trad{c.out}{\ListOf{Out}} \AND \\
  & \quad id = \Global(\B{c}) \AND ty = \llvm{"\PH{\state(\B{c.root})} *} \AND \\
  & \quad val \sep p = \IF \B{c.inst} = \text{None} \\
  & \quad \quad \THEN \llvm{"\%self\$"} \sep \llvm{""} \\
  & \quad \quad \ELSE \LET v = \variable \IN \\
  & \quad \quad \quad v \sep \\
  & \quad \quad \quad \llvm{"\PH{v} = getelementptr \PH{ \state({\B{c.inst.root}})} \%self\$, i32 0, i32 \PH{\idx(\B{c.inst})} \nl"} \\
  & \IN \\
  & \quad \llvm{"\PH{p_i}} \\
  & \quad \llvm{\PH{p_o}} \\
  & \quad \llvm{\PH{p}} \\
  & \quad \llvm{call void \PH{id}(\PH{ty}* \PH{val}, \PH{il}, \PH{ol}) \nl"}
\end{align*}
The translation of input and output arguments produces two code snippets. The
first corresponds to the LLVM instruction sequence $p$ responsible for
evaluating the argument, and the second to the type $t$ and name $v$ of the
variable storing the resulting value:
\begin{align*}
  \trad{i}{Inp} \isdef 
  & \LET p \sep v \sep t = \trad{i}{\B{Expr}} \IN p \sep \llvm{"\PH{t} \PH{v}"}\\
  \trad{o}{Out} \isdef 
  & \LET p \sep v \sep t = \trad{o}{lv} \IN p \sep \llvm{"\PH{t} \PH{v}"}
\end{align*}

@@

Initialization is done is two steps: first all imported modules are
initialized,  second the initialization clause is executed.
\begin{align*}
  \trad{i}{Init} \isdef
  & \llvm{"define void @\PH{\B{i.id}}\$init\$() \{ \nl"} \\
  & \quad\llvm{\PH{\trad{i, i.impl}{\ListOf{InitImp}}}} \\
  & \quad\llvm{\PH{\trad{i.init, "exit"}{\ListOf{\B{Inst}}}_L}} \\
  & \quad\llvm{exit: ret void \nl} \\
  & \llvm{\}"}
\end{align*}
The initialization of an imported module is performed by a call to the
corresponding initialization function:
\begin{align*}
  \lefteqn{\trad{\B{impl}, \B{impo}}{InitImp} \isdef} \\
  & \LET v = \variable \AND \B{m} = \B{impo.mach} \IN \\
  & \quad \llvm{"\PH{v} = getelementptr \PH{\state(\B{impl.root})}* \%self\$, i32 0, i32 \PH{\idx(\B{impo})}\nl}\\
  & \quad \llvm{call void @\PH{\B{m.id}}\$init\$(\PH{\state(\B{m.impl})}*
    \PH{\ell}) \nl"}
\end{align*}

@@

For each test file, indicates the B0 constructs found in the implementation.
\begin{center}
\newcommand{\no}[0]{--}
\newcommand{\yes}[0]{$\circ$}
\begin{tabular}{|l|c|c|c|c|c|c|c|c|}
\hline
Example nº & 01 & 02 & 03 & 04 & 05 & 06 & 07 & 08 \\
\hline
\hline
\multicolumn{9}{|c|}{clauses} \\
\hline
import w.o prefix & \no & \no & \no & \no & \no & \no & \yes & \no \\
import w prefix & \no & \no & \no & \no & \no & \no & \no & \yes \\
variables & \yes & \yes & \yes & \yes & \yes & \yes & \no & \yes \\
constants & \no & \yes & \no & \no & \yes & \no & \yes & \yes \\
op. input & \yes & \no & \yes & \no & \no & \no & \no & \no \\
op. output & \yes & \no & \yes & \no & \no & \no & \yes & \yes \\
\hline
\multicolumn{9}{|c|}{instructions} \\
\hline
block & \yes & \no & \yes & \yes & \no & \yes & \no & \yes \\
local var & \no & \no & \yes & \yes & \no & \yes & \yes & \yes \\
identity & \no & \no & \yes & \no & \no & \no & \no & \no \\
becomes equal & \yes & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
operation call & \no & \no & \no & \no & \no & \no & \yes & \yes \\
condition & \yes & \no & \no & \no & \yes & \no & \yes & \yes \\
case & \no & \no & \no & \no & \no & \yes & \no & \no \\
assertion & \no & \no & \no & \no & \no & \no & \no & \no \\
sequence & \yes & \yes & \yes & \yes & \no & \yes & \yes & \yes \\
while & \no & \no & \no & \yes & \no & \no & \yes & \no \\
\hline
\multicolumn{9}{|c|}{types} \\
\hline
INT & \yes & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
BOOL & \yes & \yes & \no & \no & \yes & \no & \yes & \yes \\
\hline
\multicolumn{9}{|c|}{expressions} \\
\hline
int lit & \yes & \yes & \yes & \yes & \yes & \yes & \yes & \yes \\
bool lit & \yes & \yes & \no & \yes & \yes & \no & \yes & \yes \\
vars & \yes & \yes & \yes & \yes & \no & \yes & \no & \yes \\
loc. vars & \no & \no & \yes & \yes & \no & \yes & \yes & \yes \\
const & \no & \yes & \no & \no & \yes & \no & \yes & \yes\\ 
succ & \no & \yes & \no & \yes & \yes & \no & \no & \no \\
pred & \no & \yes & \no & \yes & \yes & \no & \no & \no \\
+ & \yes & \no & \no & \yes & \no & \yes & \yes & \no \\
- & \yes & \no & \no & \no & \no & \no & \yes & \no \\
* & \no & \no & \no & \no & \no & \yes & \no & \no \\
\hline
\multicolumn{9}{|c|}{formulas} \\
\hline
and & \no & \no & \no & \no & \yes & \no & \no & \no \\
or & \no & \no & \no & \no & \yes & \no & \no & \no \\
not & \no & \no & \no & \no & \yes & \no & \no & \no \\
$<$ & \yes & \yes & \no & \no & \no & \no & \yes & \yes \\
$>$ & \no & \no & \no & \no & \yes & \no & \yes & \no \\
\hline
\end{tabular}
\end{center}
